---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "85%",
  fig.align = "center"
)
```

# CopSens

<!-- badges: start -->
<!-- badges: end -->

`CopSens` implements the copula-based sensitivity analysis method, as discussed in [Copula-based Sensitivity Analysis for Multi-Treatment CausalInference with Unobserved Confounding](https://arxiv.org/abs/2102.09412), with Gaussian copula adopted in particular.


## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("JiajingZ/CopSens")
```

The dependency `pcaMethods` from [Bioconductor](http://bioconductor.org) may fail to automatically install,
which would result in a warning similar to:
```
ERROR: dependency ‘pcaMethods’ is not available for package ‘CopSens’
```
Then, please first install the `pcaMethods` manually by 
``` r
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install("pcaMethods")
```


## Basic Usage

```{r, warning=FALSE, message=FALSE}
# load package #
library(CopSens)
```

### Example for analysis with Gaussian outcomes
```{r gaussian-outcome-example, warning=FALSE}
# load data #
y <- GaussianT_GaussianY$y
tr <- subset(GaussianT_GaussianY, select = -c(y))

# execute worst-case calibration #
est_g1 <- gcalibrate(y = y, tr = tr, t1 = tr[1:2,], t2 = tr[3:4,],
                     calitype = "worstcase", R2 = c(0.6, 1))
# visualize #
plot_estimates(est_g1)
# calculate robustness value #
rv <- cal_rv(y = y, tr = tr, t1 = tr[c(1,2,1698),], t2 = tr[c(3,4,6698),])
print(rv)

# execute multivariate calibration #
est_g2 <- gcalibrate(y = y, tr = tr, t1 = tr[1:10,], t2 = tr[11:20,],
                     calitype = "multicali", penalty_weight = c(0, 15))
# visualize #
plot_estimates(est_g2)

# execute user-specified calibration #
est_g3 <- gcalibrate(y = y, tr = tr, t1 = tr[1:2,], t2 = tr[3:4,],
                     calitype = "null", gamma = c(0.96, -0.29, 0),
                     R2 = c(0.3, 0.7, 1))
# visualize #
plot_estimates(est_g3)
# apply gamma that maximizes the bias for the first contrast considered in est_g1 #
est_g4 <- gcalibrate(y = y, tr = tr, t1 = tr[1:2,], t2 = tr[3:4,],
                     calitype = "null", gamma = est_g1$gamma[1,],
                     R2 = c(0.3, 0.7, 1))
# visualize #
plot_estimates(est_g4)
```

### Example for analysis with binary outcomes
```{r binary-outcome-example, warning=FALSE}
# load data #
y <- GaussianT_BinaryY$y
tr <- subset(GaussianT_BinaryY, select = -c(y))
t1 <- tr[1:5,]
t2 <- rep(0, times = ncol(tr))

# calibrate #
est_df <- bcalibrate(y = y, tr = tr, t = rbind(t1, t2),
                     gamma = c(1.27, -0.28, 0),
                     R2 = c(0.5, 0.7))$est_df
# calculate risk ratio estimator #
rr_df <- est_df[1:5,] / as.numeric(est_df[6,])
# visualize #
plot_estimates(rr_df)
```

## Implementation To a Mouse Obesity Study

For further illustration, we compare our approach to a recent analysis of a mouse obesity dataset [(Wang et al. (2006))](https://doi.org/10.1371/journal.pgen.0020015), conducted by [Miao et al. (2020)](https://arxiv.org/abs/2011.04504). In particular, we consider the comparison to their null treatments approach, which assumes that at least half of the confounded treatments have no causal effect on the outcome.

```{r}
# load the data #
y <- micedata[,1]
tr <- micedata[, 2:18]
```

Following Miao et al. (2020), we infer a Gaussian conditional confounder distribution by applying factor analysis to treatments, and fit the observed outcome distribution with a linear regression.

```{r}
# treatment model #
nfact <- 1
tr_factanal <- factanal(tr, factors=nfact, scores = "regression")
B_hat <- diag(sqrt(diag(var(tr)))) %*% tr_factanal$loadings
Sigma_t_u_hat <- diag(tr_factanal$uniquenesses * sqrt(diag(var(tr))))
u_hat <- tr_factanal$scores
coef_mu_u_t_hat <- t(B_hat) %*% solve(B_hat %*% t(B_hat) + Sigma_t_u_hat)
cov_u_t_hat <- diag(nfact) - t(B_hat) %*% solve(B_hat %*% t(B_hat) + Sigma_t_u_hat) %*% B_hat

# outcome model #
lmfit_y_t <- lm(y ~ ., data = micedata[,1:18])
beta_t <- coef(lmfit_y_t)[-1]
names(beta_t) <- colnames(tr)
sigma_y_t_hat <- sigma(lmfit_y_t)
```

We explore the ignorance regions for each treatment as well as causal estimates with multiple contrast criteria (MCCs) using the method described in [Zheng, D'Amour and Franks (2021)](https://arxiv.org/abs/2102.09412).

```{r}
k <- ncol(tr)
t1 <- diag(k)
t2 <- matrix(0, ncol = k, nrow = k)
u_t_diff <- (t1 - t2) %*% t(coef_mu_u_t_hat)

# worst-case calibration #
R2 <- c(0.3, 0.6, 1)
beta_cali_worstcase <- gcalibrate(y, tr, t1 = t1, t2 = t2, calitype = "worstcase",
                                  mu_y_dt = as.matrix(beta_t), sigma_y_t =  sigma_y_t_hat,
                                  mu_u_dt = u_t_diff, cov_u_t = cov_u_t_hat, R2 = R2)$est_df
rownames(beta_cali_worstcase) <- names(beta_t)

## multivariate calibration ##
# with L1 norm #
multcali_results_L1 <- gcalibrate(y, tr, t1 = t1, t2 = t2, calitype = "multicali",
                                  mu_y_dt = as.matrix(beta_t), sigma_y_t =  sigma_y_t_hat,
                                  mu_u_dt = u_t_diff, cov_u_t = cov_u_t_hat, normtype = "L1")
# with L2 norm #
multcali_results_L2 <- gcalibrate(y, tr, t1 = t1, t2 = t2, calitype = "multicali", 
                                  mu_y_dt = as.matrix(beta_t), sigma_y_t =  sigma_y_t_hat,
                                  mu_u_dt = u_t_diff, cov_u_t = cov_u_t_hat, normtype = "L2")
```

Below, we visualize the analysis results. The Spearman’s rank correlation between the estimated treatment effects of Miao et al. (2020) with the null treatment assumption ("miao$\_$nulltr") and ones by our MCC procedure with the L1 ("multicali$\_$L1") or L2 minimization ("multicali$\_$L2") are 0.90 and 0.93 respectively. In the plot below, the blue, green and yellow bars are closely grouped together for majority of treatments.

```{r, fig.width=11.8, fig.height=5.5, out.width = "100%"}
order_name <- rownames(multcali_results_L2$est_df)[order(multcali_results_L2$est_df[,2])]
summary_df <- data.frame(uncali = round(multcali_results_L2$est_df[order_name, 1], 3),
                         multicali_L1 = round(multcali_results_L1$est_df[order_name, 2], 3),
                         multicali_L2 = round(multcali_results_L2$est_df[order_name, 2], 3),
                         miao_nulltr =  mice_est_nulltr[order_name,]$esti,
                         miao_nulltr_sig = mice_est_nulltr[order_name,]$signif,
                         worstcase_lwr = beta_cali_worstcase[order_name, 'R2_1_lwr'],
                         worstcase_upr = beta_cali_worstcase[order_name, 'R2_1_upr'])
rownames(summary_df) <- order_name
plot_L1L2Null <- data.frame(summary_df[,c(1:4)], case = 1:nrow(summary_df)) %>%
  gather(key = "Type", value = "effect", - case) %>%
  ggplot() +
  ungeviz::geom_hpline(aes(x = case, y = effect, col = Type), width = 0.5, size = 1.2)  +
  scale_colour_manual(name = "",
                      values = c("#3B99B1", "#7CBA96", "#FFC300", "#F5191C"),
                      # divergingx_hcl(5, palette = "Zissou 1")[c(1, 2, 3, 5)],
                      labels = c("miao_nulltr",
                                 bquote("multicali_L1,"~R^2~"="~
                                              .(round(multcali_results_L1$R2*100,0))~"%"),
                                 bquote("multicali_L2,"~R^2~"="~
                                              .(round(multcali_results_L2$R2*100,0))~"%"),
                                 "naive")) +
  scale_x_continuous(breaks = 1:k, labels = order_name,
                     limits = c(0.5,k + 0.5)) +
  labs(y = "estimates") +
  theme_bw(base_size = 14) +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(size = 13, angle = 75, hjust = 1),
        legend.text.align = 0,
        legend.title = element_text(size=10))
print(plot_L1L2Null)
```

We also explore the worst-case ignorance region for each treatment under the Gaussian copula assumption. Even though, this assumption may not hold in practice, we can still see in the plot that 16/17 of Miao et al. (2020)'s causal estimates with null treatment assumption ("miao$\_$nulltr") are covered by our ignorance region ("worstcase $R^2 = 1$, lower"; "worstcase $R^2 = 1$, upper"). The only exception, "2010002N04Rik", whose causal effects by Miao et al. (2020) is, nevertheless, quite close to the lower bound.


```{r, fig.width=11.8, fig.height=5.5, out.width = "100%"}
bound_df <- tibble(x1 = 1:nrow(summary_df),
                 y1 = summary_df$worstcase_lwr,
                 x2 = 1:nrow(summary_df),
                 y2 = summary_df$worstcase_upr)
plot_L2NullWorst <-
  data.frame(summary_df[,c(1,3,4,6:7)], case = 1:nrow(summary_df)) %>%
    gather(key = "Type", value = "effect", - case) %>%
    ggplot() +
    ungeviz::geom_hpline(aes(x = case, y = effect, col = Type),
                         width = 0.5, size = 1, alpha = 0.8)  +
    geom_segment(data = bound_df, aes(x=x1, y=y1, xend=x2, yend=y2)) +
    scale_colour_manual(name = "",
                        values = c("#FFC300", "#F5191C", "#3B99B1", "black", "black"),
                        labels = c("miao_nulltr",
                                   bquote("multicali_L2,"~R^2~"="~
                                            .(round(multcali_results_L2$R2*100,0))~"%"),
                                   "naive",
                                   bquote("worstcase"~R^2~" = 1, lower"),
                                   bquote("worstcase"~R^2~" = 1, upper"))) +
    scale_x_continuous(breaks = 1:k, labels = order_name,
                       limits = c(0.5,k + 0.5)) +
    labs(y = "estimates") +
    theme_bw(base_size = 14) +
    theme(plot.title = element_text(hjust = 0.5),
          axis.text.x = element_text(size = 13, angle = 75, hjust = 1),
          legend.text.align = 0,
          legend.title = element_text(size=10))
print(plot_L2NullWorst)
```



