---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "85%"
)
```

# CopSens

<!-- badges: start -->
<!-- badges: end -->

`CopSens` implements a copula-based sensitivity analysis method for unobserved confounding in multi-treatment inference.


## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("JiajingZ/CopSens")
```

The dependency `pcaMethods` from [Bioconductor](http://bioconductor.org) may fail to automatically install,
which would result in a warning similar to:
```
ERROR: dependency ‘pcaMethods’ is not available for package ‘CopSens’
```
Then, please first install the `pcaMethods` manually by 
``` r
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install("pcaMethods")
```


## Basic Usage

```{r}
# load package #
library(CopSens)
```

### Example for analysis with Gaussian outcomes
```{r gaussian-outcome-example}
# load data #
y <- GaussianT_GaussianY$y
tr <- subset(GaussianT_GaussianY, select = -c(y))

# execute worst-case calibration #
est_df1 <- gcalibrate(y = y, tr = tr, t1 = tr[1,], t2 = tr[2,],
                      calitype = "worstcase", R2 = c(0.6, 1))
# visualize #
plot_estimates(est_df1)

# execute multivariate calibration #
est_df2 <- gcalibrate(y = y, tr = tr, t1 = tr[1:10,], t2 = tr[11:20,],
                      calitype = "multicali")
# visualize #
plot_estimates(est_df2)

# execute user-specified calibration #
est_df3 <- gcalibrate(y = y, tr = tr, t1 = tr[1:2,], t2 = tr[3:4,],
                      calitype = "null", gamma = c(0.96, -0.29, 0),
                      R2 = c(0.3, 0.7, 1))
# visualize #
plot_estimates(est_df3)
```

### Example for analysis with binary outcomes
```{r binary-outcome-example}
# load data #
y <- GaussianT_BinaryY$y
tr <- subset(GaussianT_BinaryY, select = -c(y))
t1 <- tr[1:5,]
t2 <- rep(0, times = ncol(tr))

# calibrate #
est_df <- bcalibrate(y = y, tr = tr, t = rbind(t1, t2),
                     gamma = c(1.27, -0.28, 0), R2 = c(0.5, 0.7))
# calculate risk ratio estimator #
rr_df <- est_df[1:5,] / as.numeric(est_df[6,])
# visualize #
plot_estimates(rr_df)
```

